Grove: A Convergent Collaborative Structure-Editor Calculus

### 2021-02-01 grove-paper.tex
updates from last two Grove meetings

% graph decomposition operators
\DeclareMathOperator{\parents}{\text{parents}}
\DeclareMathOperator{\multiParented}{\text{multiParented}}
\DeclareMathOperator{\singleParented}{\text{singleParented}}
\DeclareMathOperator{\deleted}{\text{deleted}}

@@

A graph $G : \E \rightarrow \Sigma$ is a function from edges to edge states,
where $\E = \U \times \V \times \P \times \V$,
unique IDs are drawn from some suitable set $\U$,
vertices are drawn from $\V$,
positions are drawn from $\P$,
and edge states are drawn from $\Sigma$, all defined below.

@@

Trees are often easier to work with than arbitrary graphs directly, for example
to present the graph to users or for type checking. A graph is not necessarily a
tree, but we can decompose any graph into a collection of expressions, governed
by the grammar in \autoref{fig:Base syntax}, organized into a structure we call
a grove.

A grove is a 4-tuple of the following form:
\[
  (e, MP, NP, W)
\]
The first component $e$ is the rooted expression.
The second component $MP$ is the set of expressions rooted at a vertex with multiple parents.
The third component $NP$ is the set of expressions rooted at a vertex with no parents.
The fourth component $W$ is the set of expressions derived from the remaining vertices, which must necessarily 
form cyclic structures we call wreaths, where each vertex has exactly one parent.
For example, the grove corresponding to the graph in Figure~XXX is ...
% TODO: add ref to example figure

% TODO: define helper function for finding the expression rooted at some vertex

>>>>

% graph decomposition operators
\DeclareMathOperator{\expr}{\text{expr}}
\DeclareMathOperator{\patt}{\text{patt}}
\DeclareMathOperator{\type}{\text{type}}
\DeclareMathOperator{\parents}{\text{parents}}
\DeclareMathOperator{\ancestors}{\text{ancestors}}
\DeclareMathOperator{\vertices}{\text{vertices}}
\DeclareMathOperator{\children}{\text{children}}

\DeclareMathOperator{\multiParented}{\text{multiParented}}
\DeclareMathOperator{\singleParented}{\text{singleParented}}
\DeclareMathOperator{\deleted}{\text{deleted}}

@@

A graph $G : \E \rightarrow \Sigma$ is a function from edges to edge states,
where $\E = \U \times \V \times \P \times \V$,
unique IDs are drawn from some suitable set $\U$ equipped with a total ordering,
vertices are drawn from $\V$,
positions are drawn from $\P$,
and edge states are drawn from $\Sigma$, all defined below.

@@

Trees are often easier to work with than arbitrary graphs directly, for example
to present the graph to users or for type checking. A graph is not necessarily a
tree, but we can decompose any graph into a collection of expressions, governed
by the grammar in \autoref{fig:Base syntax}, which we organize into a structure
we call a grove.

A grove is a 4-tuple of the following form:
\[
  (e_r, MP, NP, W)
\]
where
\begin{align*}
  e_r &= \expr(v_0) \\
  MP &= \{\expr(v) : |\parents(v)| > 1\} \\
  NP &= \{\expr(v) : |\parents(v)| = 0 \land v \ne v_0\} \\
  W  &= \{\expr(v) : |\parents(v)| = 1 \wedge v = \min(\ancestors(v))\}
\end{align*}

The first component $e_r$ is the expression rooted at the root vertex $v_0$.
The second component $MP$ is the set of expressions rooted at a vertex with multiple parents.
The third component $NP$ is the set of expressions rooted at a vertex with no parents.
The fourth component $W$ is the set of expressions derived from the remaining vertices, which must necessarily form cyclic structures we call wreaths, where each vertex has exactly one parent. In each of these, $\parents(v)$ is defined as follows.  \[ \parents(v) = \{v' : \exists \e = (u, v', p, v), G(\e) = \Plus\}  \] \\

For example, the grove corresponding to the graph in Figure~XXX is ...

\begin{figure}[b]
  \centering
  \begin{tikzpicture}
    \path (0cm,0cm) graph[graph style] {
      root[root style] -> {
       a/{\vWrapPlus} [> "\eWrapPlus"] -> {
        "\vSimpleTimes" [> "\eNestedPartsBob"'] -> {
         {},
         b/"\vNestedPartsAlice" [> "\eMultiParentAlice"', >merge edge]
        },
        {}
       }
      },
    };
    \path [draw,->,merge edge] (a) to ["\eMultiParentBob",out=-60,in=60] (b);
    \path (1cm,-1cm) graph[graph style] { plus/"\vSimpleTimes" };
    \path (2cm,-1cm) graph[graph style] { times/"\vMultiCycleTimes" };
    \path (2cm,-2cm) graph[graph style] { plus2/"\vSimpleTimes" };
    \path[draw,->,merge edge] (times) to ["\eMultiCycleBobTimes"'anchor=north,out=-135,in=-45] (plus);
    \path[draw,->,merge edge] (plus) to ["\eMultiCycleBobPlus"anchor=south,out=45,in=135] (times);
    \path[draw,->,merge edge] (times) to ["\eMultiCycleBobPlus"anchor=west,out=-90,in=90] (plus2);
    \path (1.5cm,-2.5cm) graph[graph style] { plus/"\vSimpleTimes" };
  \end{tikzpicture}
  \caption{Caption goes here}%
  \label{fig:Decomposition example}
\end{figure}

Figure YYY defines $\expr(v)$ wich determines the expression for a given vertex $v$.
There's one rule for each constructor form, e.g., $\ExpApp$.
In each child position, e.g., $\AppFun, \AppArg$, we call an auxiliary function $\expr'(v, p)$ which ... (explain different cases and what's going on here).

% TODO: add conflict hole to no-parent vertex

\begin{figure*}
\begin{align*}
  e_r &= \expr(v_0) \\
  MP &= \{\expr(v) : |\parents(v)| > 1\} \\
  NP &= \{\expr(v) : |\parents(v)| = 0 \land v \ne v_0\} \\
  W  &= \{\expr(v) : |\parents(v)| = 1 \wedge v = \min(\ancestors(v))\}
\end{align*}

\begin{align*}
  \parents(v) &= \{v' : \exists \e = (u, v', p, v), G(\e) = \Plus\} \\
  \ancestors(v) &= \parents(v) \cup \ancestors(\parents(v))
\end{align*}

\begin{displaymath}
\begin{array}{l}
  \expr(v=(u, Root)) = expr'(v, Root) \\
  \expr(v=(u, \ExpVar(x))) = x^u \\
  \expr(v=(u, \ExpApp)) = (\expr'(v, \AppFun)~\expr'(v, \AppArg))^u \\
  \expr(v=(u, \ExpLam)) = \lambda^u \patt'(v, \LamParam) : \type'(v, \LamType) . \expr'(v, \LamBody) \\
\end{array}
\end{displaymath}

% \begin{align*}
%   \expr(v=(u, Root)) &= expr'(v, Root) \\
%   \expr(v=(u, \ExpVar(x))) &= x^u \\
%   \expr(v=(u, \ExpApp)) &= (\expr'(v, \AppFun)~\expr'(v, \AppArg))^u \\
%   % | Exp_var of (string)
%   % | Exp_lam
%   % | Exp_app
%   % | Exp_num of (int)
%   % | Exp_plus
%   % | Exp_times
%   % | Exp_nil
%   % | Exp_cons
%   % | Exp_case
%   % (**** Pat ****)
%   % | Pat_var of (string)
%   % (**** Typ ****)
%   % | Typ_num
%   % | Typ_arrow
%   % | Typ_list
% \end{align*}

% DONE: add ref case to expr'
% TODO: point out cleverness in W set-builder def -- say that v being the min ancestor of itself means it is part of the cycle
% TODO: note to take least fixed point of ancestors

% Michael's suggestion:
%   f : A -> A U parents(v) U ancestors(v)
%   take lfp of f

% when talking about why W is defined that way: after explaining intuition, say
% ``we can formalize this intuition by taking the lfp of <the function f>.''
% fixed-point theory is part of the set of assumptions in this paper.

\begin{align*}
  \expr'(v,p) = \begin{cases}
    \hole & \children(v,p) = \varnothing \\
    \expr(v') & \children(v,p) = \{v'\} \\
    \#u & \children(v,p) = \{v'=(u,k)\} \land |\parents(v')| > 1 \\
    \#u & \children(v,p) = \{v'=(u,k)\} \land |\parents(v')| = 1 \land v' = \min(\ancestors(v')) \\
    \{\expr(v_1) | \ldots | \expr(v_n)\} & \children(v,p) = \{v_1, \ldots, v_n\} \\
  \end{cases}
\end{align*}

\begin{align*}
  \children(v, p) = \{v' : \exists \e (u, v, p, v'), G(\e) = \Plus\}
\end{align*}
\end{figure*}

% TODO: add text around these definitions -- turn them into paragraphs.
